### 2023-05-20 submission/manuscript.Rmd
Integrate figures and legends into manuscript

---
bibliography: references.bib
output:
  pdf_document:
    keep_tex: true
csl: asm.csl
geometry: margin=1.0in
header-includes:
 - \usepackage{upgreek}
 - \usepackage{booktabs}
 - \usepackage{longtable}
 - \usepackage{graphicx}
 - \usepackage{array}
 - \usepackage{multirow}
 - \usepackage{wrapfig}
 - \usepackage{float}
 - \usepackage{colortbl}
 - \usepackage{pdflscape}
 - \usepackage{tabu}
 - \usepackage{threeparttable}
 - \usepackage{threeparttablex}
 - \usepackage[normalem]{ulem}
 - \usepackage{makecell}
 - \usepackage{setspace}
 - \doublespacing
 - \usepackage[left]{lineno}
 - \linenumbers
 - \modulolinenumbers
 - \usepackage{helvet} % Helvetica font
 - \renewcommand*\familydefault{\sfdefault} % Use the sans serif version of the font
 - \usepackage[T1]{fontenc}
 - \usepackage[shortcuts]{extdash}
---

@@

# Waste not, want not: Revisiting the analysis that called the practice of rarefying microbiome data into question

@@

**Running title:** Review of "Waste not, want not"

@@

Since the development of highly parallelized sequencing technologies that enable microbiome researchers to distribute sequences across multiple samples in a single sequencing run, researchers have struggled to produce a consistent number of sequences from each sample in a dataset. It is common to observe more than 10-fold variation in the number of sequences per sample [XXXXX]. Researchers desire strategies to limit uneven sampling effort and need methods to control for he uneveness in their analyses. Of course, uneven sampling is not unique to microbiome research and is a challenge faced by all community ecologists. Common approaches to controlling uneven sampling efforts include use of proportional abundance (i.e., relative abundance), normalization of counts, parameter estimation, and rarefaction.

In 2014 Paul McMurdie and Susan Holmes published "Waste not, want not: why rarefying microbiome data is inadmissible" (WNWN) in PLOS Computational Biology [XXXXX]. This paper has had a significant impact on the approaches that microbiome researchers use to analyze 16S rRNA gene sequence data. According to Google Scholar, this paper has been cited more than 2,400 times as of April 2023. Anecdotely, I have received correspondence from researchers over the past 10 years asking how to address critiques from reviewers who criticize my correspondents' analysis for using rarefaction (e.g., see [this Twitter thread](https://twitter.com/inanna_nalytica/status/1264299305067786242)). I have also received similar comments from reviewers regarding my own work. Most recently, I received the critique for a preprint that I posted in 20XX that analyzes the practice of removing rare taxa from microbiome analyses [XXXXX]. In the process of responding to these reviewers' comments and preparing a manuscript investigating rarefaction and other approaches to control for uneven sequencing effort, I decided to reassess the WNWN study including their definitions, simulations, and analyses.

### Confusion regarding what is meant by "rarefying" and "rarefaction"

As I attempted to reproduce the results of WNWN, I noticed that the step that rarefied the data only performed one subsampling of the data. This caused me to re-inspect how McMurdie and Holmes defined "rarefying" in the following quoted text from their paper:

> Instead, microbiome analysis workflows often begin with an ad hoc library size normalization by random subsampling without replacement, or so-called rarefying [17]–[19]. There is confusion in the literature regarding terminology, and sometimes this normalization approach is conflated with a non-parametric resampling technique — called rarefaction [20], or individual-based taxon re-sampling curves [21] — that can be justified for coverage analysis or species richness estimation in some settings [21], though in other settings it can perform worse than parametric methods [22]. Here we emphasize the distinction between taxon re-sampling curves and normalization by strictly adhering to the terms rarefying or rarefied counts when referring to the normalization procedure, respecting the original definition for rarefaction. Rarefying is most often defined by the following steps [18].
>
> 1. Select a minimum library size, N~L,m~. This has also been called the rarefaction level [17], though we will not use the term here.
> 2. Discard libraries (microbiome samples) that have fewer reads than N~L,m~.
> 3. Subsample the remaining libraries without replacement such that they all have size N~L,m~.
>
> Often N~L,m~ is chosen to be equal to the size of the smallest library that is not considered defective, and the process of identifying defective samples comes with a risk of subjectivity and bias. In many cases researchers have also failed to repeat the random subsampling step (3) or record the pseudorandom number generation seed/process — both of which are essential for reproducibility.

It was unfortunate that McMurdie and Holmes used the term "rarefying" here and throughout their manuscript. The authors were correct to state that the distinction between "rarefying" and "rarefaction" is confusing and leads to their conflation. Adding to the confusion is that the papers cited in the first sentence of this quote (i.e., their references [17]–[19]) either do not use the words "rarefy" or "rarefying" or use them interchangably with "rarefaction". In my experience, subsequent researchers have conflated the results of this study of the effects of rarefying data with rarefaction of data. As an example, Willis (XXXXX) describes problems with rarefaction rather than rarefying data when citing WNWN in her paper proposing alternatives to rarefaction for use with alpha diversity data:

> Unfortunately, rarefaction is neither justifiable nor necessary, a view framed statistically by McMurdie and Holmes (2014) in the context of comparison of relative abundances.

In hindsight, as shown in the quoted text from WNWN, McMurdie and Holmes did emphasize the distinction between rarefying and rarefaction. However, because they seem to have coined a new meaning for rarefying, they added to the confusion by using the generally used verb form of rarefaction. Further confusion comes from the author's admonition in the final sentence that some researchers have failed to repeat the subsampling step. To most scientists, repeating the subsampling step is rarefaction. My preference is to use *subsampling* as the term describing the process they refer to as rarefying. In other words, subsampling is rarefaction, but with a single randomization. To minimize confusion, I will use subsamping in place of rarefying.

I propose the following definition of rarefaction:

1. Select a minimum library size, N~L,m~. Researchers are encouraged to report the value of N~L,m~.
2. Discard samples that have fewer reads than N~L,m~.
3. Subsample the remaining libraries without replacement such that they all have size N~L,m~.
4. Compute the desired metric (e.g., richness, Shannon diversity, Bray-Curtis distances) using the subsampled data
5. Repeat steps 3 and 4 a large number of iterations (e.g, 100 or 1,000). Researchers are encouraged to report the number of iterations.
6. Compute summary statistics using values generated from the subsampled data

This definition aligns well with how rarefaction was originally defined for comparing richness (i.e., the number of taxa in a community) across communities when communities are sampled to different depths [XXXXXXXXX]. With this more general approach to rarefaction, rarefaction can be performed using any alpha or beta diversity metric. This strategy has been widely used by my research group and others and is available in the mothur software package using commands such as `summary.single`, `rarefaction.single`, `dist.shared`. The procedure outlined above could also be used for hypothesis tests of differential abundance; however, thought would need to be given to how to synthesize the results of these tests across a large number of replications.

@@

### Description of "Simulation A" from WNWN

McMurdie and Holmes analyzed the effect of rarefying and other approaches on clustering accuracy using what they called "Simulation A" in their Figure 2A and elsewhere in their paper. In Simulation A, they investigated the ability to correctly assign simulated microbiome samples to one of two clusters representing two simulated treatment groups. Thankfully, the R code for Simulation A was provided by the authors in the `simulation-cluster-accuracy/simulation-cluster-accuracy-server.Rmd` R-flavored markdown file that was published as Protocol S1 in WNWN. I will outline the simulation strategy below and will reference line numbers from their R-flavored markdown file with "L" as a prefix.

For the WNWN cluster analysis, sampling distributions for the two treatment groups were generated using human fecal and ocean data originally take from the GlobalPaterns dataset (L129) [XXXXX]. To generate a fecal and ocean distribution, the authors included any operational taxonomic unit (OTU) that appeared in more than one of the 4 fecal and 3 ocean samples (L60 and L137). The OTUs were sorted by how many of the 7 samples the OTUs were observed in followed by their total abundance across all 7 samples (L139). From this sorted list they identified the identifiers of the first 2000 OTUs (L66). Returning to the 7 samples they selected the data for the corresponding 2000 OTUs and pooled the OTU abundances of the fecal and ocean samples separately to create two sampling distributions (L144, L159-160, L197-198). Next, the fecal and ocean distributions were mixed in 8 different fractions to generate two community types that differend by varying effect sizes (i.e., 1, 1.15, 1.25, 1.5, 1.75, 2, 2.5, and 3.5; L170-195, L220); an effect size of 1 generated a null model with no difference between the treatment groups. To simulate the variation in sequencing depth across the 80 samples, they normalized he number of sequences from each of the `r gp_n_samples` samples in the GlobalPatterns dataset so that the median number of sequences (&Ntilde;~L~) for the GlobalPatterns had 1,000, 2,000, 5,000, or 10,000 sequences (L324-325). They then randomly sampled the `r gp_n_samples` normalized sequencing depths to generate 80 sampling depths. From each community type, they simulated 40 samples by sampling to the desired number of sequences (L73, L230-233 and L326-327). Each simulation condition was repeated 5 times (L85). This resulted in 160 simulations (8 effect sizes x 4 median sampling depths x 5 replicates = 160 simulations). Finally, they removed rare and low prevalence OTUs in two steps. First, they removed any OTUs whose total abundance was less than 3 across all 80 samples and that did not appear in at least 3 samples (L368-386). Second, they removed any OTUs that did not have more than 1 sequence in more than 5% of the 80 samples (i.e., 4 samples) and that did not have a total abundance across the 80 samples greater than one half of the number of samples in each community type (i.e., 20) (L523-538, L551).

After generating the OTU counts for the 160 simulated communities, the authors applied several normalization methods, distance calculations, and clustering algorithms to the data. To normalize the OTU data the WNWN analysis either applied no normalization procedure, calculated OTU relative abundances, subsampled the data, performed variance stabilization using the DESeq R package, or performed Upper-Quartile log-fold change normalization using the edgeR R package. In subsampling the data, the authors either included all of the samples or removed samples whose sequencing depth fell below the 5, 10, 15, 20, 25, and 40 percentile across all 80 samples. Subsampled data were then used as input to calculate distances between samples using Bray-Curtis, Euclidean, Unweighted UniFrac, and Weighted Unifrac distances as implemented in the Phyloseq R package, Poisson distance as implemented in the PoiClaClu R package, and top-mean squared difference as implemented in the edgeR R package. Un-normalized data were used to calculate Bray-Curtis, Euclidean, Poisson, and Weighted UniFrac distances. Relative abundance data were used to calculate Bray-Curtis, Unweigthed UniFrac, and Weighted UniFrac distances. DESeq variance stabilization normalized data were used to calculate Bray-Curtis, Euclidean, and Weighted Unifrac distances. The Upper-Quartile log-fold change normalized data were only used to calculate top-mean squared difference distances. The resulting distance matrices were used to cluster the 80 samples into one of two clusters using partitioning around the medoid (PAM), K-means clustering, and hierarchical clustering. Although data for all three methods were presented in the supplementary Protocol S1, only the PAM data are presented in the main manuscript. The accuracy of the clustering assignments was quantified as the fraction of the 80 samples that were assigned to the correct cluster. Because some of the subsampling conditions removed samples those were counted as mis-clustered samples yielding minimum accuracies below 50%.

### Critique of the WNWN simulation design

@@

1. Simulated conditions were only replicated 5 times each
2. The average sizes of the libraries were small by modern standards
3. DESeq-based variance stabilization was used with distance calculation methods that are sensitive to negative values
4. A single subsampling of each dataset was evaluated rather than using rarefaction, which likely resulted in noisier data
5. Results using PAM clustering were not directly compared to those of K-means and hierarchical clustering
6. Subsampling removed the smallest 15% of the samples, which penalized accuracy values by 15 percentage points
7. The distribution of library sizes was not typical of those commonly seen in microbiome analyses
8. A filtering step was applied to remove rare taxa from the simulated datasets
9. No accounting for difference in performance when library sizes are confounded with treatment group
10. Clustering accuracy was used rather than direct comparisons of beta diversity
11. There was no consideration of effects of normalization methods on alpha diversity metrics

These points will serve as an outline for the Results section. After replicating the WNWN simulations, these points will be evaluated to reassess whether subsampling or rarefaction are "inadmissible".

@@

Before assessing the impact of the points I critiqued above, I attempted to replicate the results shown in Figures 4 and 5 of WNWN using the authors' code. I created a Conda environment that used the R version and package versions that were as close as possible to those used in WNWN. Because of the slight differences in packages, it was necessary to apply several patches to WNWN's R-flavored markdown file to render document. I was able to generate figures similar to those presented as WNWN's Figures 4 and 5. My results are shown in Figures S1 and S2 (**Figure s1_figure_3.pdf**/**Figure s1_figure_4.pdf**), respectively. The differences in results are likely due to differences in software versions and operating systems. It is also worth noting that the published versions of the two figures differ from those included in Protocol S1 within the rendered html file (`simulation-cluster-accuracy/simulation-cluster-accuracy-server.html`) and that the figure numbers are one higher in the paper than those generated by the R-flavored markdown file (i.e., Protocol S1 labels the figures as Figures 3 and 4 corresponding to the published Figures 4 and 5). Regardless of the differences, my results were qualitatively similar to that provided in WNWN's Protocol S1.

### 1. Simulated conditions were only replicated 5 times each

Each simulated condition was replicated 5 times in WNWN and the paper reports the mean and standard deviation of the replicate clustering accuracies. The relatively small number of replicates accounts for the jerkiness of the lines in WNWN's Figures 4 and 5 (e.g. the Bray-Curtis distances calculated on the DESeqVS normalized data). A better approach would have been to use 100 replicates as this would reduce the dependency of the results on the random number generator's seed. By increasing the number of replicates it was also possible to compare the probability of falsely and correctly clustering samples from the same and different treatment groups together (see points 9-11, below). Because the accuracies were not symmetric around the mean accuracy values the median and 95% confidence intervals or intraquartile range should have been reported. To test the effect of increasing the number of replicates, I pulled apart the code in `simulation-cluster-accuracy/simulation-cluster-accuracy-server.Rmd` into individual R and bash scripts that were executed using a Snakemake workflow with the same Conda environment I used above. This was necessary since the number of simulated conditions increased 20-fold with the additional replicates. Such intense data processing was not practical within a single R-flavored markdown document. Again, the observed results were qualitiatively similar to those generated using the single R-flavored markdown file (Figures S3 and S4 **pam_subsample15_fig_4.pdf**/**pam_subsample_fig_5.pdf**). The increased number of replications resulted in smoother lines and allowed me to present empirical 95% confidence intervals. For all analyses in the remainder of this paper, I used 100 randomized replicates per condition.

### 2. The average sizes of the libraries were small by modern standards

In the 10 years since WNWN was published, sequencing technology has advanced and sequence collections have grown considerably. For more modern datasets, it would be reasonable to expect a median number of sequences larger than 10,000 (see Table 1 of [Singleton Paper XXXXXXXX]). Therefore, I included an additional median depth of sampling value of 50,000 sequences with WNWN's four median sequencing sampling depths (i.e., 1,000, 2,000, 5,000, 10,000). Additional sequencing coverage would be expected to result in more robust distance values since there would be more information represented in the data. Indeed, the added sampling depth showed higher accuracy values at lower effect sizes for the combinations of normalization methods and distance calculations (Figure S3 **fig4_pam_subsample15_gp_ds.pdf**). Increased sequencing coverage also resulted in improved clustering accuracy for lower effect sizes when the library size minimum quantile was decreased (Figure S4 **fig5_pam_subsample_gp.pdf**). I will revisit the choice of the library size minimum quantile below.

```{r}
deseq_wunifrac <- cluster_data %>%
  filter(model == "gp" & distribution == "random" & fraction == 3.5 &
        filter == "filter" & transform == "deseq" & distance == "wunifrac" &
        method == "pam") %>%
  filter(n_seqs %in% c(1000, 2000, 10000)) %>%
  group_by(n_seqs) %>%
  summarize(accuracy = mean(fracCorrect),
            lci = quantile(fracCorrect, 0.025),
            uci = quantile(fracCorrect, 0.975)) %>%
  mutate(n_seqs = format(round(n_seqs), trim = TRUE, big.mark = ","))
```

### 3. DESeq-based variance stabilization was used with distance calculation methods that are sensitive to negative values

>>>>

---
bibliography: references.bib
output:
  pdf_document:
    keep_tex: true
csl: asm.csl
geometry: margin=1.0in
header-includes:
 - \usepackage{upgreek}
 - \usepackage{booktabs}
 - \usepackage{longtable}
 - \usepackage{graphicx}
 - \usepackage{array}
 - \usepackage{multirow}
 - \usepackage{wrapfig}
 - \usepackage{float}
 - \usepackage{colortbl}
 - \usepackage{pdflscape}
 - \usepackage{tabu}
 - \usepackage{threeparttable}
 - \usepackage{threeparttablex}
 - \usepackage[normalem]{ulem}
 - \usepackage{makecell}
 - \usepackage{setspace}
 - \doublespacing
 - \usepackage[left]{lineno}
 - \linenumbers
 - \modulolinenumbers
 - \usepackage{helvet} % Helvetica font
 - \renewcommand*\familydefault{\sfdefault} % Use the sans serif version of the font
 - \usepackage[T1]{fontenc}
 - \usepackage[shortcuts]{extdash}
 - \usepackage[document]{ragged2e}
---

\raggedright

@@

# Waste not, want not: Revisiting the analysis that called into question the practice of rarefaction

@@

**Running title:** Revisiting "Waste not, want not"

@@

Microbiome studies that use amplicon sequencing to characterize multiple samples use PCR to amplify 16S rRNA gene fragments using primers with distinct barcodes or index sequences [@Caporaso2010; **XXXXX**]. These barcodes allow researchers to pool PCR products and then deconvolute the resulting sequence data based on the barcode sequences. Despite their best efforts to generate equimolar pools of PCR products, it is common to observe more than 10-fold variation in the number of sequences per sample [**XXXXX**]. Researchers desire strategies to minimize uneven sequencing depth and thus need methods to control for this uneveness in their analyses. Of course, uneven sampling is not unique to microbiome research and is a challenge faced by all community ecologists [**XXXXX**]. Common approaches to controlling uneven sampling efforts include use of proportional abundance (i.e., relative abundance), scaling of counts, parameter estimation, and rarefaction [**XXXXX**].

In 2014 Paul McMurdie and Susan Holmes published "Waste not, want not: why rarefying microbiome data is inadmissible" (WNWN) in *PLOS Computational Biology* [@McMurdie2014]. Their paper has had a significant impact on the approaches that microbiome researchers use to analyze 16S rRNA gene sequence data. According to Google Scholar, this paper has been cited more than 2,400 times as of June 2023. There has been a rebuttal of WNWN that actually showed how rarefaction could be beneficial [@Weiss2017]. Anecdotely, I have received correspondence from researchers over the past 10 years asking how to address critiques from reviewers who criticize my correspondents' analysis for using rarefaction (e.g., see [this Twitter thread](https://twitter.com/inanna_nalytica/status/1264299305067786242)). I have also received similar comments from reviewers regarding my own work. Most recently, I received the critique for a preprint that I posted in 2020 that analyzed the practice of removing rare taxa from microbiome analyses [see comments linked to **XXXXX**; https://web.archive.org/web/20230308191505/http://dalmug.org/filter-review/]. In the process of responding to these reviewers' comments and preparing a manuscript investigating rarefaction and other approaches to control for uneven sequencing effort, I grew to appreaciate that there is signifcant confusion in the field over what is meant by "rarefying" and "rarefaction".

McMurdie and Holmes discussed and defined "rarefying" in WNWN (citation numbers updated from the original to those of the current study):

> "Instead, microbiome analysis workflows often begin with an ad hoc library size normalization by random subsampling without replacement, or so-called rarefying [@NavasMolina2013; @Hughes2005; @Koren2013]. There is confusion in the literature regarding terminology, and sometimes this normalization approach is conflated with a non-parametric resampling technique — called rarefaction [@Sanders1968], or individual-based taxon re-sampling curves [@Gotelli2001] — that can be justified for coverage analysis or species richness estimation in some settings [@Gotelli2001], though in other settings it can perform worse than parametric methods [@Mao2005]. Here we emphasize the distinction between taxon re-sampling curves and normalization by strictly adhering to the terms rarefying or rarefied counts when referring to the normalization procedure, respecting the original definition for rarefaction. Rarefying is most often defined by the following steps [@Hughes2005].
>
> 1. Select a minimum library size, N~L,m~. This has also been called the rarefaction level [@NavasMolina2013], though we will not use the term here.
> 2. Discard libraries (microbiome samples) that have fewer reads than N~L,m~.
> 3. Subsample the remaining libraries without replacement such that they all have size N~L,m~.
>
> Often N~L,m~ is chosen to be equal to the size of the smallest library that is not considered defective, and the process of identifying defective samples comes with a risk of subjectivity and bias. In many cases researchers have also failed to repeat the random subsampling step (3) or record the pseudorandom number generation seed/process — both of which are essential for reproducibility."

It was unfortunate that McMurdie and Holmes used the term "rarefying" here and throughout their manuscript. The authors were correct to state that the distinction between "rarefying" and "rarefaction" is confusing and leads to their conflation. Adding to the confusion is that the papers cited in the first sentence of this quote (i.e., [@NavasMolina2013; @Hughes2005; @Koren2013]) either do not use the words "rarefy" or "rarefying" or use them interchangably with "rarefaction". For example, Hughes and Hellman did not use "rarefy" and use "rarefaction" in the traditional sense with multiple subsamplings [@Hughes2005]. Meanwhile, the QIIME-based literature appears to use "rarefy" and "rarefaction" interchangibly meaning only a single subsampling ([@NavasMolina2013; @Koren2013]). Subsequent researchers have continued to conflate the results of this study of the effects of rarefying data with that of traditional rarefaction of data (**Supplemental Text**). An exemplar of the confusion is the creation of a technique that uses "repeatedly rarefying" as an approach distinct from rarefaction when they are actually the same thing [@Cameron2021].

As shown in the quoted text from WNWN, McMurdie and Holmes did emphasize the distinction between rarefying and rarefaction. However, because they seem to have coined a new definition for rarefying, they added to the confusion by using the generally used verb form of rarefaction. Further confusion comes from the author's admonition in the final sentence of the above quoted text that some researchers have failed to repeat the subsampling step. Classically, repeating the subsampling step is rarefaction [**XXXXX**]. In other words, subsampling is rarefaction, but with a single randomization. To minimize confusion, I will use "subsamping" in place of "rarefying" thorugh the remainder of this study.

For clarity, I will use the following definition of rarefaction:

1. Select a minimum library size, N~L,m~. Researchers are encouraged to report the value of N~L,m~.
2. Discard samples that have fewer reads than N~L,m~.
3. Subsample the remaining libraries without replacement such that they all have size N~L,m~.
4. Compute the desired metric (e.g., richness, Shannon diversity, Bray-Curtis distances) using the subsampled data
5. Repeat steps 3 and 4 a large number of iterations (e.g, 100 or 1,000). Researchers are encouraged to report the number of iterations.
6. Compute summary statistics (e.g., the mean) using the values from step 4.

This definition aligns with how rarefaction was originally defined for comparing richness (i.e., the number of taxa in a community) across communities when communities are sampled to different depths [**XXXXX**]. With this more general approach to the definition of rarefaction, rarefaction can be performed using any alpha or beta diversity metric. This strategy has been widely used by my research group and others and is available in the mothur software package using commands such as `summary.single`, `rarefaction.single`, `phylo.diversity`, and `dist.shared`. The procedure outlined above could also be used for hypothesis tests of differential abundance; however, consideration is needed to synthesize the results of these tests across a large number of replications.

@@

## Description and critique of "Simulation A" from WNWN

McMurdie and Holmes analyzed the effect of subsampling and other approaches on clustering accuracy using what they called "Simulation A" in their Figure 2A and elsewhere in their paper. In Simulation A, they investigated the ability to correctly assign simulated microbiome samples to one of two clusters representing two simulated treatment groups. Thankfully, the R code for Simulation A was provided by the authors in the `simulation-cluster-accuracy/simulation-cluster-accuracy-server.Rmd` R-flavored markdown file that was published as Protocol S1 in WNWN. I will outline the simulation strategy below and will reference line numbers from their R-flavored markdown file with "L" as a prefix.

For the WNWN cluster analysis, sampling distributions for the two treatment groups were generated using human fecal and ocean data originally take from the GlobalPaterns dataset (L129) [@Caporaso2010]. To generate a fecal and ocean distribution, the authors included any operational taxonomic unit (OTU) that appeared in more than one of the 4 fecal and 3 ocean samples (L60 and L137). The OTUs were sorted by how many of the 7 samples the OTUs were observed in followed by their total abundance across all 7 samples (L139). From this sorted list they identified the identifiers of the first 2,000 OTUs (L66). Returning to the 7 samples they selected the data for the corresponding 2,000 OTUs and pooled the OTU abundances of the fecal and ocean samples separately to create two sampling distributions (L144, L159-160, L197-198). Next, the fecal and ocean distributions were mixed in 8 different fractions to generate two community types that differend by varying effect sizes (i.e., 1.0, 1.15, 1.25, 1.5, 1.75, 2, 2.5, and 3.5; L170-195, L220); an effect size of 1.0 generated a null model with no difference between the treatment groups. To simulate the variation in sequencing depth across the 80 samples, they normalized the number of sequences from each of the `r gp_n_samples` samples in the GlobalPatterns dataset so that the median number of sequences (&Ntilde;~L~) across the samples had 1,000, 2,000, 5,000, or 10,000 sequences (L324-325). They then randomly sampled the `r gp_n_samples` normalized sequencing depths, with replacement, to generate 80 sampling depths. From each community type, they simulated 40 samples by sampling to the desired number of sequences (L73, L230-233 and L326-327). Each simulation condition was repeated 5 times (L85). Finally, they removed rare and low prevalence OTUs in two steps. First, they removed any OTUs whose total abundance was less than 3 across all 80 samples and that did not appear in at least 3 samples (L368-386). Second, they removed any OTUs that did not have more than 1 sequence in more than 5% of the 80 samples (i.e., 4 samples) and that did not have a total abundance across the 80 samples greater than one half of the number of samples in each community type (i.e., 20) (L523-538, L551). Simulation A consisted of 160 simulations (8 effect sizes x 4 median sampling depths x 5 replicates = 160 simulations).

After generating the OTU counts for the 160 simulations, the authors applied several normalization methods, distance calculations, and clustering algorithms to the data. To normalize the OTU data the WNWN analysis either applied no normalization procedure (L559), calculated OTU relative abundances (L392-401, L562-563), subsampled the data (L409-416, L566-569), performed variance stabilization using the DESeq R package (L478-504, L580-583), or performed Upper-Quartile log-fold change normalization using the edgeR R package (L425-457, L576-577, L622-721). In subsampling the data, the authors either included all of the samples or removed samples whose sequencing depth fell below the 5, 10, 15, 20, 25, and 40 percentile across all 80 samples (L96, L1065-1077). Non-normalized data were used to calculate Bray-Curtis, Euclidean, Poisson, and Weighted UniFrac distances (L793-798, L801-806). Relative abundance data were used to calculate Bray-Curtis, Unweigthed UniFrac, and Weighted UniFrac distances (L760-767). Subsampled data were then used as input to calculate distances between samples using Bray-Curtis, Euclidean, Unweighted UniFrac, and Weighted Unifrac distances as implemented in the Phyloseq R package, Poisson distances as implemented in the PoiClaClu R package, and top-mean squared difference as implemented in the edgeR R package (L769-775, L793-798, L801-806). DESeq variance stabilization normalized data were used to calculate Bray-Curtis, Euclidean, and Weighted Unifrac distances (L793-798). The Upper-Quartile log-fold change normalized data were only used to calculate top-mean squared difference distances (L801-806). The resulting distance matrices were used to cluster the 80 samples into one of two clusters using partitioning around the medoid (PAM), K-means clustering, and hierarchical clustering (L865-879). Although data for all three methods were presented in the supplementary Protocol S1, only the PAM data are presented in the main manuscript. The accuracy of the clustering assignments was quantified as the fraction of the 80 samples that were assigned to the correct cluster (L887-908). Since some of the subsampling conditions removed samples below a minimum sequencing depth threshold, those samples were counted as mis-clustered samples yielding minimum accuracies below 50%.

@@

1. Simulated conditions were only replicated 5 times each, potentially increasing the sensitivity of results to the choice of the random number generator seed
2. The average sizes of the libraries were small by modern standards, which may limit the generalizability of the results
3. DESeq-based variance stabilization was used with distance calculation methods that are sensitive to negative values, which likely led to nonsense distances and clusters
4. A single subsampling of each dataset was evaluated rather than using rarefaction, which likely resulted in noisier data
5. Results using PAM clustering were not directly compared to those of K-means and hierarchical clustering, although close inspection suggests that K-means may have been superior to PAM for some conditions
6. Subsampling removed the smallest 15% of the samples, which penalized accuracy values by 15 percentage points
7. The distribution of library sizes was not typical of those commonly seen in microbiome analyses, which may limit the generalizability of the results
8. A filtering step was applied to remove rare taxa from the simulated datasets, which may have skewed the shape of the community distributions
9. There was no accounting for differences in performance when library sizes are confounded with treatment group
10. Clustering accuracy was used rather than the more direct and frequently applied comparisons of beta diversity using permutation tests
11. There was no consideration of effects of normalization methods on alpha diversity metrics, which is the traditional application of rarefaction

After replicating the original WNWN simulations, these points will be evaluated to reassess whether subsampling or rarefaction are "inadmissible".

@@

Before assessing the impact of the points I critiqued above, I attempted to replicate the results shown in Figures 4 and 5 of WNWN using the authors' code. I created a Conda environment that used the R version and package versions that were as close as possible to those used in WNWN. It was necessary to patch the WNWN's R-flavored markdown file to render the document to be compatible with the overall workflow of this study. I was able to generate figures similar to those presented as WNWN's Figures 4 and 5; my results are shown in this study as Figures 1 and 2, respectively. The differences in results are likely due to differences in software versions and operating systems. It is also worth noting that the published versions of the two figures differ from those included in Protocol S1 within the rendered HTML file (`simulation-cluster-accuracy/simulation-cluster-accuracy-server.html`) and that the figure numbers are one higher in the paper than those generated by the R-flavored markdown file (i.e., Protocol S1 labels the figures as Figures 3 and 4 corresponding to the published Figures 4 and 5). Regardless of the differences, my results were qualitatively similar to that provided in WNWN's Protocol S1.

### 1. Simulated conditions were only replicated 5 times each, potentially increasing the sensitivity of results to the choice of the random number generator seed

Each simulated condition was replicated 5 times in WNWN and the paper reports the mean and standard deviation of the replicate clustering accuracies. The relatively small number of replicates accounts for the jerkiness of the lines in WNWN's Figures 4 and 5 (e.g. the Bray-Curtis distances calculated on the DESeqVS normalized data). A better approach would have been to use 100 replicates as this would reduce the dependency of the results on the random number generator's seed. By increasing the number of replicates it was also possible to compare the probability of falsely and correctly clustering samples from the same and different treatment groups together (see points 9-11, below). Because the accuracies were not symmetric around the mean accuracy values the median and 95% confidence intervals or intraquartile range should have been reported. To test the effect of increasing the number of replicates, I pulled apart the code in `simulation-cluster-accuracy/simulation-cluster-accuracy-server.Rmd` into individual R and bash scripts that were executed using a Snakemake workflow with the same Conda environment I used above. This was necessary since the number of simulated conditions increased 20-fold with the additional replicates. Such intense data processing was not practical within a single R-flavored markdown document. Again, the observed results were qualitiatively similar to those generated using the single R-flavored markdown file (Figures 3 and 4). The increased number of replications resulted in smoother lines and allowed me to present empirical 95% confidence intervals. For all analyses in the remainder of this paper, I used 100 randomized replicates per condition.

### 2. The average sizes of the libraries were small by modern standards, which may limit the generalizability of the results

In the 10 years since WNWN was published, sequencing technology has advanced and sequence collections have grown considerably. For more modern datasets, it would be reasonable to expect a median number of sequences larger than 10,000 (see Table 1 of [@Schloss2020]). Therefore, I included an additional median depth of sampling value of 50,000 sequences with WNWN's four median sequencing sampling depths (i.e., 1,000, 2,000, 5,000, 10,000). Additional sequencing coverage would be expected to result in more robust distance values since there would be more information represented in the data. Indeed, the added sampling depth showed higher accuracy values at lower effect sizes for the combinations of normalization methods and distance calculations (Figure 3). Increased sequencing coverage also resulted in improved clustering accuracy for lower effect sizes when the library size minimum quantile was decreased (Figure 4). I will revisit the choice of the library size minimum quantile below.

```{r}
deseq_wunifrac <- cluster_data %>%
  filter(model == "gp" & distribution == "random" & fraction == 3.5 &
        filter == "filter" & transform == "deseq" & distance == "wunifrac" &
        method == "pam") %>%
  filter(n_seqs %in% c(1000, 2000, 10000)) %>%
  group_by(n_seqs) %>%
  summarize(accuracy = median(fracCorrect),
            lci = quantile(fracCorrect, 0.025),
            uci = quantile(fracCorrect, 0.975)) %>%
  mutate(n_seqs = format(round(n_seqs), trim = TRUE, big.mark = ","))
```

### 3. DESeq-based variance stabilization was used with distance calculation methods that are sensitive to negative values, which likely led to nonsense distances and clusters

Close comparison of WNWN's Figure 4 and my version (Figure 3) revealed several difference between the two plots. First, in the WNWN analysis, the accuracies for the Weighted UniFrac distances at the largest effect size (i.e., 3.5) were 1.0 for median sequencing depths of `r oxford_comma(deseq_wunifrac$n_seqs)`. In my version of the analysis, the values for the same sequencing depths were `r oxford_comma(deseq_wunifrac$accuracy)`, respectively. The 95% confidence interval for these accuracies spaned between `r min(deseq_wunifrac$lci)` and `r max(deseq_wunifrac$uci)`. Second, the Bray-Curtis and Weigthed UniFrac distances calculated using DESeq normalized OTU counts were different between the original and my simultion at smaller effect sizes and had wide confidence intervals. Inspection of the DESeq normalized OTU counts revealed that the method resulted in negative values. It has been suggested that WNWW turned negative DESeq normalized counts to zero [@Weiss2017]; however, I was unable to find code in `simulation-cluster-accuracy-server.Rmd` that made this transformation. In fact, rendering the R-flavored markdown files in WNWN's Protocol S1 generated warning messages when passing the DESeq normalized counts to the Bray-Curtis calculator, which said, "results may be meaningless because data have negative entries in method 'bray'". Although the Weighted UniFrac calculator function did not generate a similar warning message, negative count values would also result in similarly meaningless distances. Both are due to the fact that the distance calculators sum the counts of each OTU in both samples being compared. In contrast, a Euclidean distance does not use a similar sum, but sums the square of the difference between the OTU abundance in each sample. Even if negative values were converted to zeroes, this would effectively the same as removing rare taxa, which could have a significant impact on the shape of the communities [@Schloss2020]. To assess the prevalence of negative counts in the simulated data, I quantified the fraction of negative values in the OTU matrix from each simulation and counted the number of simulations where the normalized OTU table had at least one negative value (Figure 5). In general the fraction of negative OTU counts increased with effect size, but decreased with sequencing effort. The fraction of simulations with at least onee negative value increased with effect size and sequencing effort. The high frequency of negative OTU counts resulted in highly variable Bray-Curtis and Weighted UniFrac values. It is likely that because the WNWN analysis only used 5 replicates that the large variation in accuracies at high effect sizes was missed initially. For the rest of this reanalysis study, I will only report results using the DESeq-based variance stabilization normalization with the Euclidean distance.
